Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-09-08 paper.tex
Replaced cache_lookup figure with pseudocode.

\usepackage{subfigure}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{multirow}
\usepackage{array}
%\usepackage{natbib}% bold math
\newcommand{\ddx}[2]{ \frac{ \partial #1}{ \partial {#2} } }
%\newcommand{\ddt}[1]{ \frac{\partial #1}{\partial t} }
\newcommand{\ddt}[1]{\partial_t #1}
\newcommand{\Grad}[1]{\nabla #1}
\newcommand{\Div}[1]{\Grad \cdot #1}
\newcommand{\Lap}[1]{\Delta #1}
\definecolor{light-gray}{gray}{0.95}

@@

@@

\begin{figure*}
\centering
\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
\caption{Cache lookup.}
\label{fig:cache_diagram}
\end{figure*}

@@

{\bf Overall execution of threshold queries}. The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
asynchronously. Each node begins evaluation of the query by first performing a cache lookup. If the data for the requested field, time-step and spatial region
are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested from
adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel. Figure \ref{fig:cache_diagram} illustrates 
this process.

\begin{figure}
\centering
\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
\label{fig:cache_overhead}
\end{figure}

\section{Experimental Results}
We evaluate the developed method for the execution of threshold queries to large numerical simulation datasets with the goal of analyzing the 
benefits and overhead from the introduction of the application-aware cache. We also analyze the scaling properties of the method. Finally, we show
that an integrated method that performs the evaluation on the database nodes near the data is several orders of magnitude faster than evaluating
a threshold locally by the user.

>>>>

\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{varwidth}
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage{multirow}
\usepackage{array}
%\usepackage{natbib}% bold math
\newcommand{\ddx}[2]{ \frac{ \partial #1}{ \partial {#2} } }
%\newcommand{\ddt}[1]{ \frac{\partial #1}{\partial t} }
\newcommand{\ddt}[1]{\partial_t #1}
\newcommand{\Grad}[1]{\nabla #1}
\newcommand{\Div}[1]{\Grad \cdot #1}
\newcommand{\Lap}[1]{\Delta #1}
\definecolor{light-gray}{gray}{0.95}

@@

\makeatletter
\def\BState{\State\hskip-\ALG@thistlm}
\makeatother

\makeatletter
\let\OldStatex\Statex
\renewcommand{\Statex}[1][3]{%
  \setlength\@tempdima{\algorithmicindent}%
  \OldStatex\hskip\dimexpr#1\@tempdima\relax}
\makeatother

@@

%\begin{figure*}
%\centering
%\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
%\caption{Cache lookup.}
%\label{fig:cache_diagram}
%\end{figure*}

@@

\begin{algorithm}
\caption{Get points above threshold using cache}\label{alg:cache_lookup}
\begin{algorithmic}[1]
\Require Dataset $d$, Field $f$, Timestep $t$, Threshold $k$, Query box $q = [x_l,y_l,z_l,x_u,y_u,z_u]$
\Procedure{GetThreshold}{}
\State $points \gets List()$
\State $updateCache \gets false$
\State $query \gets \text{SELECT * FROM cachedb..cacheInfo}$
\Statex[4] $\text{WHERE dataset = {\it d} AND field = {\it f}}$
\Statex[4] $\text{AND timestep = {\it t}}$
\State $command \gets SqlCommand(query)$
\State $reader \gets command.ExecuteReader()$
\If {$reader.HasRows()$}
\State $k_s \gets reader[``threshold"]$ \Comment{Stored threshold}
\State $start \gets reader[``startIndex"]$
\State $end \gets reader[``endIndex"]$
\State $ordinal \gets reader[``ordinal"]$
\If {$k \le k_s \And q \in \text{[start, end]}$}
\State $\textit{query} \gets \text{SELECT * FROM cachedb..cacheData}$
\Statex[4] $\text{WHERE cacheInfoOrdinal = {\it ordinal}}$
\State $command \gets SqlCommand(query)$
\State $reader \gets command.ExecuteReader()$
\While{reader.Read()}
\State $location \gets reader[``zindex"]$
\State $norm \gets reader[``dataValue"]$
\If {$norm \ge k \And location \in q$}
\State $\textit{points.Add(new Point(location, norm))}$
\EndIf
\EndWhile
\Else
\State $updateCache \gets true$
\EndIf
\Else
\State $updateCache \gets true$
\EndIf
\If {$updateCache$}
\State $\text{Retrieve data covering {\it q} from DB.}$
\ForAll{$p \in q$}
\State{Compute $f$ at $p$.}
\If{$\| f(p) \| \ge k$}
\State $\textit{points.Add(new Point(p, $\| f(p) \|$))}$
\EndIf
\EndFor
\State{Update cacheInfo and cacheData tables.}
\EndIf
\State{\Return $points$}
\EndProcedure
\end{algorithmic}
\end{algorithm}

The entire cache is comprised of two database tables. The {\it cacheInfo} table stores metadata for the cached entries. It stores information about the
dataset, field, time-step, start and end coordinates of the spatial region examined and the threshold value used. The {\it cacheData} table stores the locations
of all of the grid points, for which the field queried has a norm higher than the specified threshold. The {\it cacheData} table is foreign key constrained with
the ordinal of the {\it cacheInfo} table. This allows us to quickly find a record in the {\it cacheInfo} table and retrieve all of the cached entries using an index
lookup.

{\bf Overall execution of threshold queries}. Algorithm \ref{alg:cache_lookup} illustrates the process of obtaining all points with norms of the specified field
above the given threshold from the database in the presence of a cache. We described the overall execution of threshold queries to the JHTDB below. 
The mediator submits a query to each of the database nodes storing the raw data
asynchronously. Each node begins evaluation of the query by executing Algorithm \ref{alg:cache_lookup}. First a cache lookup is performed.
If the data for the requested field, time-step and spatial region
are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
parameters the query needs to be evaluated from the raw data. In those cases the raw data are read into memory with data along the boundary requested from
adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel.

\section{Experimental Results}
We evaluate the developed method for the execution of threshold queries to large numerical simulation datasets with the goal of analyzing the 
benefits and overhead from the introduction of the application-aware cache. We also analyze the scaling properties of the method. Finally, we show
that an integrated method that performs the evaluation on the database nodes near the data is several orders of magnitude faster than evaluating
a threshold locally by the user.
\begin{figure}
\centering
\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
\label{fig:cache_overhead}
\end{figure}
